---
title: "Research Question 3"
---

**Research Question 3:**

Is there a distinct pattern or clustering of individuals based on their health-related attributes and behaviours?

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Load necessary libraries
library(ggplot2)
library(corrplot)

# Read the data
health_data <- read.csv("C:\\Users\\arava\\OneDrive\\Desktop\\STAT515\\modified_obesity_data.csv")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Convert columns to appropriate data types if needed
health_data$Age <- as.numeric(as.character(health_data$Age))
health_data$Height <- as.numeric(as.character(health_data$Height))
health_data$Weight <- as.numeric(as.character(health_data$Weight))
# Convert other columns to appropriate data types as needed...

# Summary statistics
summary(health_data)

# Histograms
par(mfrow=c(3, 3))
for (i in 2:ncol(health_data)) {
  if (is.numeric(health_data[, i])) {
    hist(health_data[,i], main=names(health_data)[i], xlab="")
  }
}

# Box plots
par(mfrow=c(3, 3))
for (i in 2:ncol(health_data)) {
  if (is.numeric(health_data[, i])) {
    boxplot(health_data[,i] ~ health_data$Obesity_Levels, main=names(health_data)[i], xlab="Obesity Levels", ylab=names(health_data)[i])
  }
}

# Scatter plot matrix
# Filter numeric columns
numeric_cols <- sapply(health_data, is.numeric)
numeric_data <- health_data[, numeric_cols]

# Scatter plot matrix
pairs(numeric_data, pch = 16)

```

This code is designed for exploratory data analysis (EDA) of a dataset related to health and obesity. It aims to provide insights into the characteristics of the dataset, including summary statistics, distributions of variables, relationships between variables, and potential patterns or trends.

First, we will load two libraries - ggplot 2 for creating graphical visualizations and corrplot for plotting the correlation matrices. 

Then we will read the dataset from a csv file named obesity data and we have stored it in a variable named health data. 

Convert specific columns such as age, weight, and height to numeric data types.  It first coerces them into character type and then converts them into numeric type. This is often necessary when R reads these values as factors or characters instead of numeric.

Now let's get the summary statistics for the dataset including minimum, maximum, mean, median, 1st quartile and 3rd quartile values for the numeric columns. 

Create histograms for each numeric variable in the dataset. It plots them in a 3x3 grid using the par function, iterating over each numeric column, and plotting a histogram for each one. 

Create the box plots for each numeric variable grouped by 'Obesity Levels'.  It also arranges them in a 3x3 grid similar to the histograms. 

Create a scatter plot matrix for all the numeric variables in the dataset. It first filters out only numeric columns, stores them in numeric data, and then creates a scatter plot matrix using the pairs function, with points represented by filled circles (pch = 16).

```{r}
#| code-fold: true
#| code-summary: "Show the code"
health_data <- read.csv("C:\\Users\\arava\\OneDrive\\Desktop\\STAT515\\modified_obesity_data.csv")

# Select relevant features
selected_features <- health_data[, c("Age", "Height", "Weight", "family_history_with_overweight", "Highcaloric_food", 
                                     "Vegetable_intake", "Main_Meal_Intake", "Food_between_meals", "Smoking_Habit", 
                                     "Water_intake", "Calorie_Intake", "PHYSICAL_Activity", "Technology_dependency", 
                                     "Alcohol_Consumption", "Mode_of_Transportation")]

# Check for missing values and handle if necessary
# For example: selected_features <- na.omit(selected_features)

# Perform feature scaling or normalization
# Here, we'll use z-score standardization
# Exclude non-numeric columns before scaling
numeric_features <- selected_features[, sapply(selected_features, is.numeric)]

# Perform feature scaling
scaled_features <- scale(numeric_features)

```

The purpose of this code is to prepare the dataset for further analysis or modeling by selecting relevant features, handling missing values if necessary, and standardizing numeric features through z-score standardization. These preprocessing steps are essential for ensuring the quality and consistency of the data before performing any downstream tasks such as modeling or statistical analysis.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
library(cluster)
library(factoextra)

# Read the data
health_data <- read.csv("C:\\Users\\arava\\OneDrive\\Desktop\\STAT515\\modified_obesity_data.csv")  

# Select relevant features for clustering
selected_features <- health_data[, c("Age", "Height", "Weight", "family_history_with_overweight", "Highcaloric_food", 
                                     "Vegetable_intake", "Main_Meal_Intake", "Food_between_meals", "Smoking_Habit", 
                                     "Water_intake", "Calorie_Intake", "PHYSICAL_Activity", "Technology_dependency", 
                                     "Alcohol_Consumption", "Mode_of_Transportation")]


# Convert categorical variables to factors if needed
# Convert categorical variables to dummy variables
selected_features <- model.matrix(~ . - 1, data = selected_features)

# Perform K-means clustering
k=3
kmeans_result <- kmeans(selected_features, centers = k)


# Handle missing values if any
# Example: selected_features <- na.omit(selected_features)

# Perform K-means clustering
k <- 3  # Number of clusters
set.seed(123)  # Set seed for reproducibility
kmeans_result <- kmeans(selected_features, centers = k)


# Adjust plot parameters for better visibility
fviz_cluster(kmeans_result, 
             geom = "point", 
             data = selected_features, 
             stand = FALSE,  # Disable scaling for better visibility
             ellipse.type = "convex",  # Use convex hulls for clusters
             ellipse.level = 0.95,  # Confidence level for ellipses
             palette = "jco",  # Color palette
             ggtheme = theme_minimal() +  # Combine themes
               theme(legend.position = "bottom"),  # Adjust legend position
             main = paste("K-means Clustering (K =", k, ")"),  # Main title
             show.clust.cent = TRUE,  # Show cluster centers
             pointsize = 2)  # Adjust point size for better visibility

# Evaluate clusters
silhouette_score <- silhouette(kmeans_result$cluster, dist(selected_features))

```

This segment performs k-means clustering on the health and obesity dataset, visualizes the resulting clusters, and evaluates their quality using the silhouette score. Here is the break down of step by step process.

Firstly, we will import the required libraries for performing clustering (cluster) and visualizing clustering results (factoextra).

Read the dataset from a CSV file and stores it in the variable health_data.

Select specific features from the dataset that are deemed relevant for clustering. These features will be used as input variables for the k-means clustering algorithm

Convert categorical variables into dummy variables, which are required for clustering algorithms like k-means that operate on numerical data.

Perform k-means clustering on the selected features. It specifies the number of clusters (k), sets a seed for reproducibility, and then executes the k-means algorithm.

Visualize the clusters obtained from the k-means algorithm. It plots the data points, cluster centers, and cluster ellipses for better interpretation of the clusters.

Calculate the silhouette score to evaluate the quality of the clusters. The silhouette score provides a measure of how well-separated the clusters are and helps in assessing the effectiveness of the clustering algorithm.

**Cluster Interpretation:**

**Based on the clustering results, we can observe distinct groups of individuals with different health profiles. The clusters can be interpreted as follows:**

**Cluster 1:**

Individuals with relatively lower age, height, and weight, with mixed dietary habits and moderate physical activity levels.

**Cluster 2:**

Individuals with higher age, height, and weight, often having family history with overweight, consuming high-caloric food frequently, and exhibiting low physical activity levels.

**Cluster 3:**

Individuals with moderate age and height, but higher weight, frequently having family history with overweight, consuming high-caloric food frequently, and exhibiting high physical activity levels.

**Conclusion:**

The clustering analysis revealed distinct groups of individuals with varying health attributes and behaviors. Understanding these clusters can inform personalized interventions and public health policies aimed at promoting healthy lifestyles and preventing overweight and obesity.
